Community Detection Methods for Generalized Random Dot Product Graphs

### 2021-06-01 proposal/proposal.Rmd
updated slides
fixed typo

Since the communities are subspaces in the latent space, we can also use existing methods for subspace clustering on the ASE. Of particular interest is Sparse Subspace Clustering (SSC), which is performed by solving an optimization problem for each observed point in a sample. Given $X \in \mathbb{R}^{n \times d}$ with vectors $x_i^\top \in \mathbb{R}^d$ as rows of $X$, the optimization problem $c_i = \arg\min\limits_{c} \|c\|_1$ subject to $x_i = X_{-i} c$ and $c^{(i)} = 0$ is solved for each $i \in [n]$. The solutions are collected into matrix $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top$ to construct affinity matrix $B = |C| + |C^\top|$. If each $x_i$ lie perfectly on one of $K$ subspaces, $B$ is sparse such that $B_{ij} = 0$ $\forall x_i, x_j$ belonging to different subspaces. Then $B$ can describe a graph with at least $K$ disjoint subgraphs, and if the number of subgraphs is exactly $K$, each subgraph maps onto a subspace.

@@

\begin{equation} \label{eq:ssc}
c_i = \arg\min_c \frac{1}{2} \|x_i - X_{-i} c \|_2^2 + \lambda \|c\|_1
\end{equation}

@@

# Estimated Timeline of Completion

>>>>

Since the communities are subspaces in the latent space, we can also use existing methods for subspace clustering on the ASE. Of particular interest is Sparse Subspace Clustering (SSC), which is performed by solving an optimization problem for each observed point in a sample. Given $X \in \mathbb{R}^{n \times d}$ with vectors $x_i^\top \in \mathbb{R}^d$ as rows of $X$, the optimization problem $c_i = \arg\min\limits_{c} \|c\|_1$ subject to $x_i = X_{-i}^\top c$ and $c^{(i)} = 0$ is solved for each $i \in [n]$. The solutions are collected into matrix $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top$ to construct affinity matrix $B = |C| + |C^\top|$. If each $x_i$ lie perfectly on one of $K$ subspaces, $B$ is sparse such that $B_{ij} = 0$ $\forall x_i, x_j$ belonging to different subspaces. Then $B$ can describe a graph with at least $K$ disjoint subgraphs, and if the number of subgraphs is exactly $K$, each subgraph maps onto a subspace.

@@

\begin{equation} \label{eq:ssc}
c_i = \arg\min_c \frac{1}{2} \|x_i - X_{-i}^\top c \|_2^2 + \lambda \|c\|_1
\end{equation}

@@

## Estimated Timeline of Completion

### 2021-06-01 proposal/slides.Rmd
fixed typo

## Connecting the PABM to the GRDPG ($K = 2$)

**Theorem** (KTT): 
Given popularity vectors $\{\lambda^{(kl)}\}_2$, we can define block diagonal $X \in \mathbb{R}^{n \times 4}$ and $U \in \mathbb{O}(4)$ such that $A \sim \PABM(\{\lambda^{(kl)}\}_2)$ is equivalent to $A \sim \GRDPG_{3, 1}(X U)$.

Proof: Decompose $P$ as follows

$$X = \begin{bmatrix}
\lambda^{(11)} & \lambda^{(12)} & 0 & 0 \\
0 & 0 & \lambda^{(21)} & \lambda^{(22)}
\end{bmatrix}$$

$$U = \begin{bmatrix}
1 & 0 & 0 & 0 \\
0 & 0 & 1 / \sqrt{2} & 1 / \sqrt{2} \\
0 & 0 & 1 / \sqrt{2} & -1 / \sqrt{2} \\
0 & 1 & 0 & 0
\end{bmatrix}$$

$$P = (X U) I_{3, 1} (X U)^\top = 
\begin{bmatrix}
\lambda^{(11)} (\lambda^{(11)})^\top & \lambda^{(12)} (\lambda^{(21)})^\top \\
\lambda^{(21)} (\lambda^{(12)})^\top & \lambda^{(22)} (\lambda^{(22)})^\top
\end{bmatrix}$$

@@

$$\Lambda^{(k)} = \begin{bmatrix} 
\lambda^{(k1)} & \cdots & \lambda^{(kK)} 
\end{bmatrix}$$

@@

1. Solve $n$ optimization problems $c_i = \arg\min_c \|c\|_1$ 
subject to $x_i = X c$ and $c^{(i)} = 0$.  
This is typically performed via LASSO: 
$c_i = \arg\min \frac{1}{2} \|x_i - X_{-i} c\|_2^2 + \lambda \|c\|_1$
2. Compile solutions $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}$.
3. Construct affinity matrix $B = |C| + |C^\top|$.

@@

If $X$ obeys the *Subspace Detection Property*, then $B$ is sparse such that 
$B_{ij} = 0$ if $x_i$ and $x_j$ belong to different subspaces.

@@

```{r, fig.height = 5, fig.width = 5, out.width = '75%'}
set.seed(314159)

@@

```{r, fig.height = 5, fig.width = 5, out.width = '75%'}
Y <- t(X)
Y <- sweep(Y, 2, apply(Y, 2, function(y) sqrt(sum(y ** 2))), `/`)
Y[is.nan(Y)] <- 0
lambda <- .05
N <- ncol(Y)
B <- plyr::aaply(seq(N), 1, function(i) {
  y <- Y[, i]
  X <- Y[, -i]
  betahat <- glmnet::glmnet(X, y, lambda = lambda, intercept = FALSE) %>% 
    coef() %>% 
    as.numeric()
  if (i != N) {
    betahat <- c(betahat[seq(i)], 0, betahat[seq(i + 1, N)])
  } else {
    betahat <- c(betahat, 0)
  }
  betahat <- betahat[-1]
  return(betahat)
}, .parallel = TRUE) %>% 
  abs()
B <- sweep(B, 2, apply(B, 2, max), `/`)
B[is.nan(B)] <- 0
W <- B + t(B)

>>>>

@@

$$X = \begin{bmatrix}
\Lambda^{(1)} & \cdots & 0 \\
0 & \ddots & 0 \\
0 & \cdots & \Lambda^{(K)}
\end{bmatrix} 
\in \mathbb{R}^{n \times K^2}$$

$$\Lambda^{(k)} = \begin{bmatrix} 
\lambda^{(k1)} & \cdots & \lambda^{(kK)} 
\end{bmatrix} 
\in \mathbb{R}^{n_k \times K}$$

@@

1. Solve $n$ optimization problems $c_i = \arg\min_c \|c\|_1$ 
subject to $x_i = X^\top c$ and $c^{(i)} = 0$.  
This is typically performed via LASSO: 
$c_i = \arg\min \frac{1}{2} \|x_i - X_{-i}^\top c\|_2^2 + \lambda \|c\|_1$
2. Compile solutions $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}$.
3. Construct affinity matrix $B = |C| + |C^\top|$.

@@

If $X$ obeys the *Subspace Detection Property*, then $B$ is sparse such that 
$B_{ij} = 0$ for each $x_i$ and $x_j$ belonging to different subspaces.

@@

```{r, fig.height = 3, fig.width = 4, out.width = '50%'}
set.seed(314159)

par(mar = rep(1, 4))

@@

```{r, fig.height = 3, fig.width = 5, out.width = '75%'}
Y <- t(X)
Y <- sweep(Y, 2, apply(Y, 2, function(y) sqrt(sum(y ** 2))), `/`)
Y[is.nan(Y)] <- 0
lambda <- .05
N <- ncol(Y)
B <- plyr::aaply(seq(N), 1, function(i) {
  y <- Y[, i]
  X <- Y[, -i]
  betahat <- glmnet::glmnet(X, y, lambda = lambda, intercept = FALSE) %>% 
    coef() %>% 
    as.numeric()
  if (i != N) {
    betahat <- c(betahat[seq(i)], 0, betahat[seq(i + 1, N)])
  } else {
    betahat <- c(betahat, 0)
  }
  betahat <- betahat[-1]
  return(betahat)
}, .parallel = TRUE) %>% 
  abs()
B <- sweep(B, 2, apply(B, 2, max), `/`)
B[is.nan(B)] <- 0
W <- B + t(B)
